---
title: "RakuNotes"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---
```{r echo = FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

# Chapter 9 Arrays and List

The values in a list or in an array are called *elements* or *items*.

1. List: **Ordered** and **Immutable**
2. Arrays: **Variables** and generally **Mutable**

## List
```{r eval=FALSE}
> 3,4,5
(3 4 5)
> say (3, 4, 5).WHAT
(List)
> say (1..10).WHAT
(Range)
```

## Arrays
```{r}
> my @odd_digits = 1, 3, 5, 7, 9;
[1 3 5 7 9]
> my @single_digit_numbers = 0..9;
[0 1 2 3 4 5 6 7 8 9]
> my @weekdays = <mon tue wed thu fri>;
[mon tue wed thu fri]
```

Arrays can be **heterogenous**.
```{r}
> my @heterogenous-array = 1, 2.3, pi, "str", (1, 2, 4);
[1 2.3 3.14159265358979 str (1 2 4)]
```

Unassigned arrays are empty arrays.
```{r}
> my @empty;
[]
> my @empty = (); # Explicitly declared empty arrays
[]
```

Arrays are accessed using `@array[i]` syntax.
```{r}
> say "The third day is @weekdays[2]";
The third day is wed
> my @even-digits = 0, 2, 4, 6, 8;
[0, 2, 4, 6, 8]
> my @small-even_digits = @even-digits[0..2]
[0 2 4]
> my @min-max-even-digits = @even-digits[0, 4]
[0 8]
```

Arrays will expand if the index of assignment correspond to items that does not exist.
```{r}
my @arrays = 1, 2, 3;
@arrays[7] = 9;
say @arrays;
# [1 2 3 (Any) (Any) (Any) (Any) 9]
```


### Useful Methods

#### Adding and Removing Elements
1. `shift` Removes and returns first elements.
2. `pop` Removes and returns last elements.
3. `unshift` Adds an item or a list of items to the beginning of array.
4. `push` Adds an item or a list of items to the end of an array

_`push` and `unshift` **does not** flatten the elements inserted by default._

Use `|` prefix operator to flatten the added array.
```{r}
> my @numbers = <2 4 6 7>;
[2 4 6 7]
> @numbers.push(<8 10>);
[2 4 6 7 (8 10)]
> @numbers.push(|<8 10>);
[2 4 6 7 (8 10) 8 10]
```

The `:delete` adverb can be used to delete(and return) any elements in an array.
The deleted elements are replaced with undefined item `(Any)`.
```{r}
my @fruit = <apple banana pear cherry pineapple orange>;
my $removed = @fruit[2]:delete;
say $removed; # pear
say @fruit; [apple banana (Any) cherry pineapple orange]
```


#### Array Length
```{r}
> say @heterogenous-array.elems; # Return Array length
5
```

#### Identical Arrays
```{r}
# eqv
# To compare structure, type and value
my @even = 0, 2, 4, 6, 8;
my @even2 = reverse 8, 6, 4, 2, 0;
say @even eqv @even2; # True

# Smart match ~~
# To compare content only
say <1 2 3 4 5> eqv 1..5; # False
say <1 2 3 4 5> ~~ 1..5; #True

# Numeric equality ==
# Compare array length; == coerces arguments to numeric type
say <1 2 3 4 5> == <5 6 7 8 9>; # True
```

#### Reverse Slicing
```{r}
my @reverse-small-even_digits = @even-digits[reverse 0..2];
my @reverse-small-even_digits = reverse @even-digits[0..2];
```

#### Splicing
`splice` can add, remove, and return one or several items to or from an array.
```{r}
my @out_array = splice @array, $start, $num_elems, @replacement; 
# @replacement is optional
```

#### Traversing List
##### `for` loop
```{r}
my @colors = <red orange yellow green blue indigo violet>;
for @colors -> $color {
  say $color;
}
```

##### Read and re-assign elements (doubly pointy block)

Works on array but not list (immutable)
```{r}
my @colors = <red orange yellow green blue indigo violet>;
for @colors <-> $color {
  $color = tc $color; # turn every element Title Case
}
```

Using topical variable `$_`
```{r}
for colors {
  $_ = $_.uc
}
```

##### By index
```{r}
for 0..@colors.end -> $idx {
  say "$idx @colors[$idx]";
}
```

##### By Keys

`keys` will return a list of indices for the array or the list
```{r}
for keys @colors -> $idx {
  say "$idx @colors[$idx]";
}
```

##### By Keys-Values
`kv` returns an interleaved sequence of indexes and values
```{r}
for @letters.kv -> $idx, $val {
  say "$idx, $val";
}

my @letters = 'a'..'e';
say @letters.kv; # (0 a 1 b 2 c 3 d 4 e)
```

##### `loop`
C-like loop
```{r}
loop (my $i = 0; $i < 5; $i++){
  say $i, "->" ~ @letters[$i];
}
```

Infinite Loop
```{r}
loop{
  # do something
  #last if...
}
```


#### Unique Elements
```{r}
> say <a b d c a f d g>.unique;
(a b d c f g)
```

Use `squish` if the array is known to be **sorted**.

### Placeholders

`^` is a twigil, which state that the variable is a **placeholder parameter** or a **self declared positional parameter**. The variables are associated to the input according to their **lexicographic order**.
```{r}
my @letters = 'a'..'e';
for @letters.kv{
  say "$^a -> $^b"; # 'a' comes before 'b' hence first input is given to $a
}
```

## Stacks and Queues

1. `stack` is a last in/ first out (LIFO) data structure
2. `queue` is a first in/ first out (FIFO) data structure
